## Quick start

```{r setup}
library(varmapack)
library(randompack)
```

Create a model from coefficient matrices and its innovation covariance. A
single lag may be supplied as a matrix; multiple lags use an `r` by `r` by lag
array.

```{r varma}
A <- matrix(c(0.5, 0.1,
              0.0, 0.3), 2, 2)
B <- matrix(c(0.2, 0.0,
              0.1, 0.1), 2, 2)
model <- varmapack_model(A = A, B = B, Sig = diag(2))

rng <- randompack_rng()
rng$seed(123)
X <- model$sim(100, nrep = 3, rng = rng)
dim(X)
```

The first dimension is the series dimension, the second is time, and the third
selects the replicate. Requesting shocks returns a named list.

```{r shocks}
out <- model$sim(20, nrep = 2, rng = rng, return_shocks = TRUE)
names(out)
dim(out$E)
```

Built-in testcases return model objects.

```{r testcases}
varmapack_testcases()
test_model <- varmapack_testcase("smallARMA1")
test_model$specrad()
```

Model methods provide theoretical autocovariances, impulse responses, and the
AR and MA spectral radii.

```{r analysis}
Gamma <- model$acvf(10)
Psi <- model$psi(10)
Theta <- model$irf(10)
model$specrad()
model$ma_specrad()
```

The `varmapack_autocov()` function computes sample autocovariances for an
observed time-series matrix with variables in rows.

```{r autocov}
varmapack_autocov(X[, , 1], maxlag = 5)
```

VARMAX models use exogenous coefficient matrices `C` and input values `z`.
They require fixed starting values `X0`.

```{r varmax}
C <- array(c(0.3, -0.2), c(2, 1, 1))
varmax <- varmapack_model(A = A, B = B, C = C, Sig = diag(2))
X0 <- matrix(0, 2, 2)
z <- matrix(sin(seq_len(100)/10), 1, 100)
Xmax <- varmax$sim(100, X0 = X0, z = z, rng = rng)
```
