---
title: "Varmapack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Varmapack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Varmapack simulates Gaussian vector autoregressive-moving-average models with
exact stationary initialization. The first returned values therefore have the
model-implied distribution without discarding a burn-in segment.

```{r setup}
library(varmapack)
library(randompack)
```

Create a VARMA model from coefficient matrices and its innovation covariance.
Matrices with one lag may be supplied directly; arrays use dimensions `r`, `r`,
and lag.

```{r varma}
A <- matrix(c(0.5, 0.1,
              0.0, 0.3), 2, 2)
B <- matrix(c(0.2, 0.0,
              0.1, 0.1), 2, 2)
model <- varmapack_model(A = A, B = B, Sig = diag(2))

rng <- randompack_rng()
rng$seed(123)
X <- model$sim(100, nrep = 3, rng = rng)
dim(X)
```

The first dimension is the series dimension, the second is time, and the third
selects the replicate. Requesting shocks returns a named list.

```{r shocks}
out <- model$sim(20, nrep = 2, rng = rng, return_shocks = TRUE)
names(out)
dim(out$E)
```

Varmapack also supplies theoretical autocovariances, impulse responses, and
spectral radii directly from the model.

```{r analysis}
Gamma <- model$acvf(10)
Psi <- model$psi(10)
Theta <- model$irf(10)
model$specrad()
model$ma_specrad()
```

VARMAX models use `C` and the corresponding exogenous input sequence `z`.
They require explicitly supplied starting values.

```{r varmax}
C <- array(c(0.3, -0.2), c(2, 1, 1))
varmax <- varmapack_model(A = A, B = B, C = C, Sig = diag(2))
X0 <- matrix(0, 2, 2)
z <- matrix(sin(seq_len(100)/10), 1, 100)
Xmax <- varmax$sim(100, X0 = X0, z = z, rng = rng)
```
